---
title: "Data Analysis and Prediction for Life Expectancy"
author: "Group 106: Artsiom Strok, Lin Jiang, Mayank Chhablani"
date: 'July 15, 2019'
output:
  html_document:
    toc: yes
  pdf_document: default
urlcolor: cyan
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
cache = TRUE
```

## Introduction

***

### Team information

Name               NetID     Email
----------------   -------   --------------------
Artsiom Strok      astrok2   astrok2@illinois.edu
Lin Jiang          linj3     linj3@illinois.edu
Mayank Chhablani   mchhab2	 mchhab2@illinois.edu
----------------   -------   --------------------

### Description of the dataset

Life expectancy has increased dramatically over the last few centuries. Since 1900 the global average life expectancy has more than doubled and there has been a huge development in health sector in the past 15 years resulting in improvement of human mortality rates especially in the developing nations in comparison to the past 30 years. Therefore, in this project, only data from year 2000-2015 is considered for exploration and analysis. 

This dataset is a compilation of data from the Global Health Observatory (GHO) and United Nations. The GHO data repository is WHO's gateway to health-related statistics which provides access to a variety of indicators on priority health topics including mortality and burden of diseases, environmental health, violence and injuries etc. (http://apps.who.int/gho/data/node.resources). The economic data such as GDP is collected from the National Accounts Main Aggregates Database under United Nations which collects and disseminates economic statistics from countries worldwide (https://unstats.un.org/unsd/snaama/Index).

This dataset is cleaned by removing some missing values, maily for population, Hepatitis B and GDP from less known countries and shared on Kaggle website (https://www.kaggle.com/kumarajarshi/life-expectancy-who). The final dataset contains 2938 observations with 22 variables which are more critical and representative among all the categories of health-related factors from year 2000 - 2015 for 193 countries.

The description of each variable for this dataset is listed below:

**Numerical Response**

- `Life expectancy`: Life Expectancy in age

**Numerical Predictors**

- `Year`: Year
- `Adult Mortality`: Adult Mortality Rates of both sexes (probability of dying between 15 and 60 years per 1000 population)
- `infant`: deathsNumber of Infant Deaths per 1000 population
- `Alcohol`: Alcohol, recorded per capita (15+) consumption (in litres of pure alcohol)
- `percentage expenditure`: Expenditure on health as a percentage of Gross Domestic Product per capita(%)
- `Hepatitis B`: Hepatitis B (HepB) immunization coverage among 1-year-olds (%)
- `Measles`: Measles - number of reported cases per 1000 population
- `BMI`: Average Body Mass Index of entire population
- `under-five deaths`: Number of under-five deaths per 1000 population
- `Polio`: Polio (Pol3) immunization coverage among 1-year-olds (%)
- `Total expenditure`: General government expenditure on health as a percentage of total government expenditure (%)
- `Diphtheria`: Diphtheria tetanus toxoid and pertussis (DTP3) immunization coverage among 1-year-olds (%)
- `HIV/AIDS`: Deaths per 1 000 live births HIV/AIDS (0-4 years)
- `GDP`: Gross Domestic Product per capita (in USD)
- `Population`: Population of the country
- `thinness 1-19 years`: Prevalence of thinness among children and adolescents for Age 10 to 19 (% )
- `thinness 5-9 years`: Prevalence of thinness among children for Age 5 to 9(%)
- `Income composition of resources`: Human Development Index in terms of income composition of resources (index ranging from 0 to 1)
- `Schooling`: Number of years of Schooling(years)

**Categotical Predictors**

- `Country`: Country
- `Status`: Developed or Developing status

### Background information

The GHO data repository is WHO's gateway to health-related statistics which provides access to a variety of indicators on priority health topics including mortality and burden of diseases, environmental health, violence and injuries etc. (http://apps.who.int/gho/data/node.resources).

The economic data such as GDP is collected from the National Accounts Main Aggregates Database under United Nations which collects and disseminates economic statistics from countries worldwide (https://unstats.un.org/unsd/snaama/Index).


By combining the data from these two databases and exploring the interactions between the variables such as immunization, moratlity, education and economic factors, we expect to build a robust multiple linear regression model to predict the life expectancy more accurately.


### Statement of interest

For this project, we would like to:

- identify the major factors that affect the life expectancy
- explore the interactions between the variables such as immunization, moratlity, education, economic factors etc.
- discover the corelation between different factors
- perform residual and outlier diagnostics to exclude some influential points if any
- build a robust multiple linear regression model to predict the life expectancy

### Samples of the data
```{r message=FALSE, warning=FALSE, cache = cache}

library(knitr)
data = read.csv("Life Expectancy Data.csv")
kable(t(data[sample(nrow(data), 5), ]))
```


***

## Methods

### Required packages:

- DAAG
- corrplot
- ggplot2
- magrittr
- knitr
- tidyr
- GGally
- faraway
- lmtest
- MASS


```{r, message=FALSE, warning=FALSE, include=FALSE, cache = cache}
library(DAAG)
library(corrplot)
library(ggplot2)
library(magrittr)
library(knitr)
library(tidyr)
library(GGally)
library(faraway)
library(lmtest)
library(MASS)
set.seed(106)
```

```{r warning=FALSE, include=FALSE, cache = cache}
calculate_rmse <- function(actual, predicted) {
  mean((actual - predicted) ^ 2)
}

#BP-Test: Equal variance test
calculate_bptest <- function(model) {
  bptest(model)$p.value
}

#Shaipro-Test: Normality test
calculate_shapiro_test <- function(model) {
  shapiro.test(resid(model))$p.value
}

#Adjusted R_squared test
calculate_adj_r2 <- function(model) {
  summary(model)$adj.r.squared
}

#LOOCV
calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

#Get metrics in table
show_metrics <- function(model) {
  kable(
    data.frame(
      bptest = toString(unname(calculate_bptest(model))),
      shapiro_test = toString(unname(calculate_shapiro_test(model))),
      adj_r2 = toString(calculate_adj_r2(model)),
      LOOCV = toString(calc_loocv_rmse(model)),
      TRAIN_RMSE = toString(calculate_rmse(data.train$Life.expectancy, predict(model, data.train))),
      TEST_RMSE = toString(calculate_rmse(data.test$Life.expectancy, predict(model, data.test)))
    )
  )
}
```



### Data analysis:

```{r, cache = cache}
original_data = read.csv("Life Expectancy Data.csv")
kable(t(original_data[sample(nrow(original_data), 5), ]))
kable(sort(colSums(is.na(original_data)), decreasing = TRUE), col.names = "Number of missing values")
```


1289 samples have at least one missing value. Alcohol is missing in 194 samples all of them belongs to 2015 year and these countiries definately should have alcohol consuption more than 0. The data was collected in 2015 when data about alcohol consumption simply was not available.

Life expextancy and adult mortality is missing for 10 samples in 2013, all of them belongs to islands.

Hepatitis B is missing in 553 samples. Samples belongs to different countries and years.

```{r, cache = cache}
nrow(original_data)
data = na.omit(original_data)
nrow(original_data) - nrow(data)
data = data[-1] #exclude country from dataset
kable(t(do.call(cbind, lapply(data, summary))))
```


Looking at the summary data we can already see some inconsistencies. In `Infant Deaths` we see that the max value listed is 1600 which doesn't make sense since we're working with per 1000 population data. The same or similar numbers we can see for `Infant deaths`, `Measles`, `Under five deaths`

```{r, cache = cache}
boxplot(data$Adult.Mortality)
boxplot(data$infant.deaths)
boxplot(data$Measles)
boxplot(data$under.five.deaths)

data = data[
  data$Adult.Mortality < boxplot(data$Adult.Mortality, plot = FALSE)$stats[5] &
    data$infant.deaths < boxplot(data$infant.deaths, plot = FALSE)$stats[5] &
    data$Measles < boxplot(data$Measles, plot = FALSE)$stats[5] &
    data$under.five.deaths < boxplot(data$under.five.deaths, plot = FALSE)$stats[5]
  , ]
nrow(data)
```

This original dataset contains `Country` variable which we would not use for model building. Therefore, this column is excluded from the following data analysis as well as the records with `NA` in some of the columns mentioned above. This will reduce the size of the original dataset from 2938 to 1649 rows, which captures majority of the information and allows the speed of modeling to be more efficient.


### Data Visualization

Let's take a look on histograms for each attribute in the dataset.

```{r, cache = cache}
ggplot(gather(data[-2]), aes(value)) +
    geom_histogram(bins = 10) + 
    facet_wrap(~key, scales = 'free_x')
```

Let's take a look of the dataset using plots of `Life Expectancy` vs `Status` or `Year`. The boxplot indicates that there is significant difference in `Life Expectancy` between the `Developed` and `Developing` countries. As expected, the `Life Expectancy` increases as the `Year` passes by and the Violin plot shows that the data is well distributed across different years.

```{r, cache = cache}
par(mfrow = c(1, 2))
# Histogram of Life Expectancy
hist(
  data$Life.expectancy,
  xlab = "Life Expectancy",
  main = "Distribution of Life Expectancy",
  col = "dodgerblue",
  breaks = 25
)
# Boxplot of Life Expectancy vs Country Status (Developed vs Developing)
plot(
  data$Status,
  data$Life.expectancy,
  xlab = "Status",
  ylab = "Life Expectancy",
  main = "Life Expectancy vs. Status",
  col = c(2, 3)
)

# Violin plot of Life Expectancy vs. Year
data %>% ggplot() + geom_violin(aes(
  x = Year,
  y = Life.expectancy,
  group = Year,
  fill = Year
)) + labs(title = "Life Expectancy vs. Year")
```


### Colinearity

Colinearity issue is visualized using the following plots. Resulsts show that some predictors have strong collinearity issues, such as `infant.deaths` vs. `under.five.deaths`, `GDP` vs. `percentage.expenditure`, `Population` vs. `thinness..1.19.years` etc.

```{r fig.height=15, fig.width=15, message=FALSE, warning=FALSE, cache=cache}
corrplot(
  cor(data[-c(2)]),
  method = "color",
  order = "hclust",
  type = "lower",
  outline = TRUE,
  tl.col = "black",
  addCoef.col = "darkgreen",
)
ggpairs(data[-c(2)])
```

### Train test split


```{r message=FALSE, warning=FALSE, cache=cache}
# train test split 70/30 hold out
train_size = floor(0.7 * nrow(data))
train_idx = sample(nrow(data), train_size)
data.train = data[train_idx, ]
data.test = data[-train_idx, ]
```

### Model Building

`Steps to generate best model:`

We will start with simple additive model and then use AIC/ BIC to weed out unnecessary predictors. We then, may try out interactions  and see if that improves our model and later, we may use response or, predictor transformations to improve our selected model.

`Additve Model with all predictors`

```{r, cache = cache}
model.additive = lm(Life.expectancy ~ ., data.train)
#summary(model.additive)
par(mfrow = c(2, 2))
plot(model.additive)
```

So, looking at `Residual vs Fitted plot` that we may have issue of linearity and equal-variance. Similary, we need to check for `normality assumption` as dictated by `Normal Q-Q plot`

Let's check for `Homoscedasticity`and `Normality` assumptions.

```{r, cache = cache}
show_metrics(model.additive)
```

So, from above data it looks liek although `Adjusted R^2` and `RMSE` is better, but, we `REJECT` both `Homoscedasticity`and `Normality` assumptions. Also, from summary of model it seems there is need to prune out some predictors.

Let's apply AIC and BIC to see if we can improve our model

```{r, cache = cache}
model.additive.step.aic = step(model.additive, trace = FALSE)

par(mfrow = c(2, 2))
plot(model.additive.step.aic)
```

So, looking at AIC model, we can see that `GDP`vs `percentage.expenditure` collinearity issue has been removed by AIC. Also, some of the non-significant predictors are also removed.

```{r, cache = cache}
show_metrics(model.additive.step.aic)
```

Apart from reduction in predictors, there isn't a significant improvement. Let's try out BIC:

```{r, cache = cache}
model.additive.step.bic = step(model.additive, k = log(nrow(data.train)), trace = FALSE)

par(mfrow = c(2, 2))
plot(model.additive.step.bic)
```


```{r, cache = cache}
show_metrics(model.additive.step.bic)
```

So, BIC and AIC resutls are almost same and so does the evaluatin metric but, we saw two improvements: (1) `Reduction in number of predictors` (2) `Reduction in influential points` as per the `Residual vs Leverage ` graph.

Hence, let's select AIC model and try interactions. The BIC is little aggressive in removing predictors so, we are selecting AIC additive model.

Also, let's check for variance-inflation-factor

```{r, warning=FALSE, cache = cache}
vif_aic_all_model = vif(model.additive.step.aic)
vif_aic_all_model[vif_aic_all_model > 5]
```

As evident from above `VIF` computation, `infant.deaths` and `under.five.deaths` in the model are causing `collinearity issue`. This same issue we found during our correlation plot. Let's try to improve the model and remove issues which are violating linearity assumptions.

`Applying Two-Way Interaction`:


```{r, cache = cache}
model.interaction = lm(Life.expectancy ~ . ^ 2, data = data.train)
```

Since, this will explode the number of predictors let's generate AIC and BIC models for this interaction model.

AIC Interaction Model:

```{r, cache = cache}
model.interaction.step.aic = step(model.interaction, trace = FALSE)

par(mfrow = c(2, 2))
plot(model.interaction.step.aic)
```

```{r, cache = cache}
show_metrics(model.interaction.step.aic)
```

AIC and BIC model are again somewhat similar.
Total numbr of predictors used by AIC is 119 and we can see that with interactions we have made improvement in `adjusted r_squared` and `bptest`. Let's see the `VIF imapct`


```{r, cache = cache}
vif_interaction_model = vif(model.interaction.step.aic)
vif_interaction_model[vif_interaction_model > 5]
```


This interaction model is getting overly complicated. Let's inspect pairs scatter plot.


```{r fig.height=30, fig.width=30, cache = cache}
pairs(data.frame(data$Life.expectancy, data$Year, data$Status, data$Adult.Mortality, data$infant.deaths, data$Alcohol, data$percentage.expenditure, data$BMI, data$under.five.deaths, data$Polio, data$Total.expenditure, data$HIV.AIDS, data$thinness.5.9.years, data$Income.composition.of.resources, data$Schooling))
```

Based on this cross-predictor plots, we confirm that (a) `HIV.AIDS` has non-linear relationship with `Life.expectancy' and, (B) `HIV.AIDS` and `Adult.Mortality` are also interacting non-linearly. Also, we may condsider removing `Year` because it may divert model in different direction because of its chaotic relationship with the response.

Check for impact of high VIF predictors.

```{r, cache = cache}
model_selected_wo_underfive = lm(
  Life.expectancy ~ Year + Status + Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + BMI + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling,
  data = data.train
)

model_selected_wo_infant.deaths = lm(
  Life.expectancy ~ Year + Status + Adult.Mortality + under.five.deaths + Alcohol + percentage.expenditure + BMI + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling,
  data = data.train
)

model_selected_underfive = lm(
  under.five.deaths ~ Year + Status + Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + BMI  + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling,
  data = data.train
)

model_selected_infant.deaths = lm(
  infant.deaths ~ Year + Status + Adult.Mortality + under.five.deaths + Alcohol + percentage.expenditure + BMI + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling,
  data = data.train
)

cor(resid(model_selected_underfive),
    resid(model_selected_wo_underfive))

cor(resid(model_selected_infant.deaths),
    resid(model_selected_wo_infant.deaths))

```


Based on this we can get rid of `under.five.deaths`


```{r, cache = cache}
model_selected = lm(
  Life.expectancy ~ Year + Status + Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + BMI   + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling ,
  data = data.train
)
par(mfrow = c(2, 2))
plot(model_selected)
```

```{r, cache = cache}
show_metrics(model_selected)
```

```{r, cache = cache}
vif(model_selected)
```

Looks like none of our predictor is causing inflation in variance.
Let's consider interaction of these predictors.


```{r, cache = cache}
model_selected_interactions = lm(
  Life.expectancy ~ (
    Year + Status + Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + BMI   + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling
  ) ^ 2 ,
  data = data.train
)
par(mfrow = c(2, 2))
plot(model_selected_interactions)
```

```{r, cache = cache}
show_metrics(model_selected_interactions)
```

```{r, cache = cache}
model_selected_interactions_aic = step(model_selected_interactions, trace = FALSE)
#summary(model_selected_interactions_aic)
```

```{r, cache = cache}
show_metrics(model_selected_interactions_aic)
```

```{r, cache = cache}
model_selected_interactions_bic=step(model_selected_interactions, trace=0, k = log(nrow(data.train)))
#summary(model_selected_interactions_bic)

show_metrics(model_selected_interactions_bic)
par(mfrow=c(2,2))
plot(model_selected_interactions_bic)

```

```{r, cache = cache}
model_selected_poly_interactions=lm(Life.expectancy ~ Year + Status + Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + BMI   + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling + infant.deaths:Polio +  Adult.Mortality:HIV.AIDS + I(HIV.AIDS^2) , data=data.train)
par(mfrow = c(2, 2))
plot(model_selected_poly_interactions)
show_metrics(model_selected_poly_interactions)
```

```{r, cache = cache}
model_selected_poly_interactions=lm(Life.expectancy ~ Year + Status + Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + BMI   + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling + Adult.Mortality:HIV.AIDS +I(HIV.AIDS^2) , data=data.train)
par(mfrow = c(2, 2))
plot(model_selected_poly_interactions)
show_metrics(model_selected_poly_interactions)
vif_poly = vif(model_selected_poly_interactions)
vif_poly[vif_poly > 5]
length(coef(model_selected_poly_interactions))
```

```{r, cache = cache}
model_selected_poly_interactions=lm(Life.expectancy ~ Year + Status + Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + BMI   + Polio + Total.expenditure + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling + Adult.Mortality:HIV.AIDS +I(HIV.AIDS^2) , data=data.train)
model_selected_poly_interactions_aic=step(model_selected_poly_interactions, trace=0, k=log(nrow(data)))
par(mfrow = c(2, 2))
plot(model_selected_poly_interactions_aic)
show_metrics(model_selected_poly_interactions_aic)
vif_poly = vif(model_selected_poly_interactions_aic)
vif_poly[vif_poly > 5]
length(coef(model_selected_poly_interactions_aic))
```

```{r, cache = cache}
model_selected_outliers = lm(
  Life.expectancy ~  Adult.Mortality  +
    infant.deaths + Polio + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources  +
    Schooling  + Adult.Mortality:HIV.AIDS + I(HIV.AIDS ^ 2),
  data = data.train,
  subset = abs(rstandard(model_selected_poly_interactions_aic)) < 2
)

par(mfrow = c(2, 2))
plot(model_selected_outliers)
show_metrics(model_selected_outliers)
```


Based on above plots, we can say that we have achieved a good model based with `10` predictors and the formula is:

`formula = Life.expectancy ~ Adult.Mortality + infant.deaths +
    Polio + HIV.AIDS + thinness.5.9.years + Income.composition.of.resources +
    Schooling + Adult.Mortality:HIV.AIDS + I(HIV.AIDS^2)`

Also, looking at the model it seems, we are satisfying `assumptions of linear model`.

Evaluation metric are as follows:

```{r, cache = cache}
show_metrics(model_selected_outliers)
```

Box-Cox shows that $\lambda$ = 0.5 is in the confidence interval and close to the maxinum, which suggests a tranformation of the form.
```{r, cache = cache}
# Box-Cox Transformation
par(mfrow = c(1, 2))
boxcox(model_selected_outliers, plotit = TRUE)
boxcox(model_selected_outliers,
       plotit = TRUE,
       lambda = seq(0, 1, by = 0.1))
```

Explore the relationship between Life.expectancy and each predictor.
```{r, cache = cache}
plot(Life.expectancy ~ Adult.Mortality, data = data.train, col = "grey", pch = 20, cex = 1.5,
main = "Life.expectancy vs Adult.Mortality")

plot(Life.expectancy ~ infant.deaths, data = data.train, col = "grey", pch = 20, cex = 1.5,
main = "Life.expectancy vs infant.deaths")

plot(Life.expectancy ~ Polio, data = data.train, col = "grey", pch = 20, cex = 1.5,
main = "Life.expectancy vs Polio")

plot(Life.expectancy ~ HIV.AIDS, data = data.train, col = "grey", pch = 20, cex = 1.5,
main = "Life.expectancy vs HIV.AIDS")

plot(Life.expectancy ~ thinness.5.9.years, data = data.train, col = "grey", pch = 20, cex = 1.5,
main = "Life.expectancy vs thinness.5.9.years")

plot(Life.expectancy ~ Income.composition.of.resources, data = data.train, col = "grey", pch = 20, cex = 1.5,
main = "Life.expectancy vs Income.composition.of.resources")

plot(Life.expectancy ~ Schooling, data = data.train, col = "grey", pch = 20, cex = 1.5,
main = "Life.expectancy vs Schooling")
```

```{r, cache = cache}
lm_le_am = lm(Life.expectancy ~ Adult.Mortality, data = data.train)
show_metrics(lm_le_am)
lm_le_am_log = lm(Life.expectancy ~ log(Adult.Mortality), data = data.train)
show_metrics(lm_le_am_log)
# bp.test result is significantly improved with log transformation of Adult.Mortality.
lm_le_am_poly = lm(Life.expectancy ~ Adult.Mortality + I(Adult.Mortality ^ 2), data = data.train)
show_metrics(lm_le_am_poly)
# Polynomial transformation doesn't work.
```


```{r, cache = cache}
lm_le_id = lm(Life.expectancy ~ infant.deaths, data = data.train)
show_metrics(lm_le_id)
# lm_le_id_log = lm(Life.expectancy ~ log(infant.deaths), data = data.train)
# couldn't perform log transformation because of O values.

lm_le_id_poly = lm(Life.expectancy ~ infant.deaths + I(infant.deaths ^ 2), data = data.train)
show_metrics(lm_le_id_poly)
# Polynomial transformation of infant.deaths doesn't improve.
```

```{r, cache = cache}
lm_le_po = lm(Life.expectancy ~ Polio, data = data.train)
show_metrics(lm_le_po)
lm_le_po_log = lm(Life.expectancy ~ log(Polio), data = data.train)
show_metrics(lm_le_po_log)
# bp.test result is improved with log transformation of Polio.
lm_le_po_poly = lm(Life.expectancy ~ Polio + I(Polio ^ 2), data = data.train)
show_metrics(lm_le_po_poly)
# Polynomial transformation also helps.

```

```{r, cache = cache}
lm_le_hiv = lm(Life.expectancy ~ HIV.AIDS, data = data.train)
show_metrics(lm_le_hiv)
lm_le_hiv_log = lm(Life.expectancy ~ log(HIV.AIDS), data = data.train)
show_metrics(lm_le_hiv_log)
# bp.test result is significantly improved with log transformation of HIV.AIDS.
lm_le_hiv_poly = lm(Life.expectancy ~ HIV.AIDS + I(HIV.AIDS ^ 2), data = data.train)
show_metrics(lm_le_hiv_poly)
# Polynomial transformation also helps.
lm_le_hiv_log_poly = lm(Life.expectancy ~ log(HIV.AIDS) + I(HIV.AIDS ^ 2), data = data.train) # doesn't further improve
show_metrics(lm_le_hiv_log_poly)
```

```{r, cache = cache}
lm_le_thn = lm(Life.expectancy ~ thinness.5.9.years, data = data.train)
show_metrics(lm_le_thn)
lm_le_thn_log = lm(Life.expectancy ~ log(thinness.5.9.years), data = data.train)
show_metrics(lm_le_thn_log)
# bp.test result is significantly improved with log transformation.
lm_le_thn_poly = lm(Life.expectancy ~ thinness.5.9.years + I(thinness.5.9.years ^ 2),
                    data = data.train)
show_metrics(lm_le_thn_poly)
# Polynomial transformation doesn't help a lot.
```


```{r, cache = cache}
lm_le_icr = lm(Life.expectancy ~ Income.composition.of.resources, data = data.train)
show_metrics(lm_le_icr)
# lm_le_icr_log = lm(Life.expectancy ~ log(Income.composition.of.resources), data = data.train) couldn't perform log tranformation due to 0 values.

# bp.test result is significantly improved with log transformation.
lm_le_icr_poly = lm(
  Life.expectancy ~ Income.composition.of.resources + I(Income.composition.of.resources ^ 2),
  data = data.train
)
show_metrics(lm_le_icr_poly)
# Polynomial transformation helps a little.
```

```{r, cache = cache}
lm_le_sch = lm(Life.expectancy ~ Schooling, data = data.train)
show_metrics(lm_le_sch)
lm_le_sch_log = lm(Life.expectancy ~ log(Schooling), data = data.train)
show_metrics(lm_le_sch_log)
# log transformation doesn't work.
lm_le_sch_poly = lm(Life.expectancy ~ Schooling + I(Schooling ^ 2), data = data.train)
show_metrics(lm_le_sch_poly)
# Polynomial transformation doesn't help very much.
```



```{r, cache = cache}
# New model based on the transformation result
model_trans = lm(Life.expectancy ~ log(Adult.Mortality) + infant.deaths + Polio  + I(Polio ^ 2) + HIV.AIDS + I(HIV.AIDS^2) + log(thinness.5.9.years) + Income.composition.of.resources + Schooling + Adult.Mortality:HIV.AIDS, data = data.train)
show_metrics(model_selected_outliers)
show_metrics(model_trans)

# After transformation, it shows that bptest and shapiro test results improves a little.
```


## Results

The best model stastified all LINE assumptions. To build the model we performed data analysis where we removed all records with `NA` values and outliers. We looked at correlation matrix between predictors and excluded highly correlated predictors. We started from simple additive model then we tried to search better model using backward search using both AIC and BIC evaluation criteria. Next we tried to apply 2 level interactions for every predictor and perform search again. We pick the best selected model and started manually explore different interactions and transformations. 

```{r, cache = cache}
best_model = model_selected_interactions_bic
par(mfrow = c(2, 2))
plot(best_model)
show_metrics(best_model)
```

## Discussion

This dataset is quite complex since it contains 22 variables of health-related factors from year 2000 - 2015 for 193 countries. We didn't group data by country because we would like to build a more general model for prediction, however, this increases the challenge for model building. By ploting the response vs the single predictor, such as Life.expectancy vs Adult.Mortality, we can see that some countries has quite different distribution of Adult.Mortality and this will weaken the robustness of the final model by combining data from different countries together.

Eventually we build a quite satisfying multiple linear regression model using AIC & BIC evaluation, dummy variables exploration, variable interaction, residual and outlier diagnoistics, response transformation, predictor polynomial regression, etc. This model has high bptest value and adjusted R square, low LOOCV and RMSE values. The shapiro test result is lower than 0.01 but the normal Q-Q plot looks good.